### 2015-06-07 chapters/pitfalls.tex
updated pitfall absence equivalence

The character, represented here by glyph <Å>, is encoded in the Unicode Standard
in the first two examples by a single-character sequence; each is assigned a
different code point. In the third example, the glyph is encoded in a
multiple-character sequence that is composed of two character code points. All
three sequences are \textsc{canonically equivalent}, i.e.~they are strings that
represent the same abstract character and because they are not distinguishable
by the user, the Unicode Standard requires them to be treated the same in
regards to their behavior and appearance. Nevertheless, they are encoded
differently. For example, if one were to search an electronic text (with
software that does not apply Unicode Standard normalization) for
\textsc{angstrom sign} (\uni{212B}), instances of \textsc{latin capital letter a
with ring above} (\uni{00C5}) would not be found.

@@

As illustrated, some characters in the Unicode Standard have alternative
representations (in fact, many do), but the Unicode Normalization Algorithm can
be used to transform certain sequences of characters into canonical (or
compatible) forms to test for equivalency. To determine equivalence, each
character in the Unicode Standard is associated with a combining class, which is
formally defined as a character property called \textsc{canonical combining
class} which is specified in the Unicode Character Database. The combining class
assigned to each code point is a numeric value between 0 and 254 and is used by
the Unicode Canonical Ordering Algorithm to determine which sequences of
combining marks are canonically equivalent. Normalization forms, as very briefly
described above, can be used to ensure character equivalence by ordering
character sequences so that they can be faithfully compared.

It is very important to note that any software applications that is Unicode
Standard compliant is free to change the character stream from one
representation to another. This means that a software application may compose,
decompose or reorder characters as its developers desire; as long as the
resultant strings are canonically equivalent to the original. This might lead to
unexpected behavior for users. Various players, like the Unicode Consortium, the
W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and some
software applications that we tested indeed seem to automatically convert
strings into NFC.\footnote{See the summary of various recommendation here:
\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.}
This means in practice that if a user, for example, enters <a> and < ̀>,
i.e.~\textsc{latin small letter a} at \uni{0061} and \textsc{combining grave
accent} at \uni{0300}, this might be converted into <à>, i.e.~\textsc{latin
small letter a with grave} at \uni{00E0}.\footnote{The behavior of software
applications can be quite erratic in this respect. For example, Apple's TextEdit
does not do any conversion on text entry. However, when you copy and paste some
text inside the same document in rich text mode (i.e.~RTF-format), it
will be transformed into NFC on paste. Saving a document does not do any
conversion to the glyphs on screen, but it will save the characters in NFC.}

@@

Although in most cases canonical equivalence will take care of alternative
encodings of the same character, there are a some cases in which the Unicode
Standard decided against equivalence. This leads to identical characters that
are not equivalent, like <ø> \textsc{latin small letter o with stroke} at
\uni{00F8} and <o̷> a combination of \textsc{latin small letter o} at \uni{006F}
with \textsc{combining short solidus overlay} at \uni{0037}.
The general rule followed is that extensions of Latin characters that are
connected to the character are not separated as combining diacritics. For
example, characters like <ƞ ɲ ɳ> or <ɰ ɱ> are obviously derived from <n> and <m>
respectively, but they are treated like new separate characters in the Unicode
Standard. Likewise, characters like <ø> and <ƈ> are not separated into a base 
character <o> and <c> with an attached combining diacritic.

Interestingly, and somewhat illogically, there are three elements attached to
other characters that are still treated as separable in the Unicode Standard. First,
the cedilla, like in <ç> \textsc{latin small letter c with cedilla} at
\uni{00E7}, second the ogonek, like in <ą> \textsc{latin small letter a with
ogonek} at \uni{0105} and, third, the horn, like in <ơ> \textsc{latin small
letter o with horn} at \uni{01A1}. These characters are indeed decomposed in a 
base character with a combining diacritic. However, it is these cases that should be 
considered the exceptions to the rule.

There are two combinations that deserve specific care because it is actually 
possible to produce identical characters in different ways, without them being 
canonical equivalent. First, there are the
combining characters designated as ``overlay'' in the Unicode Standard, like    
<\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334} or 
<\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}. 
There are many characters that look like they are 
precomposed with such an overlay, for example 
<\charis{ɫ ᵬ ᵭ ᵱ}> or 
<\charis{ƚ ɨ ɉ ɍ}>, or also the 
example of <ø> given at the start of this section. Second, the same situation also 
occurs with combining characters designated as ``hook'', like < {\large  ̡}> 
\textsc{combining palatalized hook below} at \uni{0321}. This element seems to
occur in precomposed characters like <\charis{ᶀ ᶁ ᶂ ᶄ}>. 
However, all these characters are not canonical equivalent with a combination of 
their base characters with the combining overlay or combining hook.

>>>>

The character, represented here by glyph <Å>, is encoded in the Unicode Standard
in the first two examples by a single-character sequence; each is assigned a
different code point. In the third example, the glyph is encoded in a
multiple-character sequence that is composed of two character code points. All
three sequences are \textsc{canonically equivalent}, i.e.~they are strings that
represent the same abstract character and because they are not distinguishable
by the user, the Unicode Standard requires them to be treated the same in
regards to their behavior and appearance. Nevertheless, they are encoded
differently. For example, if one were to search an electronic text (with
software that does not apply Unicode Standard normalization) for
\textsc{angstrom sign} (\uni{212B}), then the instances of \textsc{latin 
capital letter a with ring above} (\uni{00C5}) would not be found.

@@

As illustrated, some characters in the Unicode Standard have alternative
representations (in fact, many do), but the Unicode Normalization Algorithm can
be used to transform certain sequences of characters into canonical
forms to test for equivalency. To determine equivalence, each
character in the Unicode Standard is associated with a combining class, which is
formally defined as a character property called \textsc{canonical combining
class} which is specified in the Unicode Character Database. The combining class
assigned to each code point is a numeric value between 0 and 254 and is used by
the Unicode Canonical Ordering Algorithm to determine which sequences of
characters are canonically equivalent. Normalization forms, as very briefly
described above, can be used to ensure character equivalence by ordering
character sequences so that they can be faithfully compared.

It is very important to note that any software applications that is Unicode
Standard compliant is free to change the character stream from one
representation to another. This means that a software application may compose,
decompose or reorder characters as its developers desire; as long as the
resultant strings are canonically equivalent to the original. This might lead to
unexpected behavior for users. Various players, like the Unicode Consortium, the
W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and
some software applications that we tested indeed seem to automatically convert
strings into NFC.\footnote{See the summary of various recommendation here:
\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
practice that if a user, for example, enters <a> and < ̀>, i.e.~\textsc{latin
small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
this might be automatically converted into <à>, i.e.~\textsc{latin small letter
a with grave} at \uni{00E0}.\footnote{The behavior of software applications can
be quite erratic in this respect. For example, Apple's TextEdit does not do any
conversion on text entry. However, when you copy and paste some text inside the
same document in rich text mode (i.e.~RTF-format), it will be transformed into
NFC on paste. Saving a document does not do any conversion to the glyphs on
screen, but it will save the characters in NFC.}

@@

Although in most cases canonical equivalence will take care of alternative
encodings of the same character, there are a some cases in which the Unicode
Standard decided against equivalence. This leads to identical characters that
are not equivalent, like <ø> \textsc{latin small letter o with stroke} at
\uni{00F8} and <o̷> a combination of \textsc{latin small letter o} at \uni{006F}
with \textsc{combining short solidus overlay} at \uni{0037}.
The general rule followed is that extensions of Latin characters that are
connected to the base character are not separated as combining diacritics. For
example, characters like <ŋ ɲ ɳ> or <ɖ ɗ> are obviously derived from <n> and <d>
respectively, but they are treated like new separate characters in the Unicode
Standard. Likewise, characters like <ø> and <ƈ> are not separated into a base 
character <o> and <c> with an attached combining diacritic.

Interestingly, and somewhat illogically, there are three elements, which are
directly attached to their base characters, but which are still treated as
separable in the Unicode Standard. Such characters are decomposed (in NFD
normalization) in a base character with a combining diacritic. However, it is
these cases that should be considered the exceptions to the rule. These three 
elements are the following:

\begin{itemize}

  \item the \textsc{combining cedilla} at \uni{0327} \newline This diacritic is
        for example attested in the precomposed character <ç> \textsc{latin
        small letter c with cedilla} at \uni{00E7}. This <ç> will thus be
        decomposed in NFC normalization.
  \item the \textsc{combining ogonek} at \uni{0328} \newline This diacritic is
        for example attested in precomposed <ą> \textsc{latin small letter a
        with ogonek} at \uni{0105}. This <ą> will thus be decomposed in NFC
        normalization.
  \item the \textsc{combining horn} at \uni{031B} \newline This diacritic is for
        example attested in precomposed <ơ> \textsc{latin small letter o with
        horn} at \uni{01A1}. This <ơ> will thus be decomposed in NFC
        normalization.

\end{itemize}

There are further combinations that deserve specific care because it is actually
possible to produce identical characters in different ways without them being
canonically equivalent. In these situations, the general rule holds, namely that
characters with attached extras are not decomposed. However, in the following
cases the ``extras'' actually exist as combining diacritics, so there is also 
the possibility to construct a character by using a base character with those 
combining diacritics.

\begin{itemize}

  \item First, there are the combining characters designated as ``combining
        overlay'' in the Unicode Standard, like <\ \ {\large ̴}>
        \textsc{combining tilde overlay} at \uni{0334} or <\ \ {\large ̵}>
        \textsc{combining short stroke overlay} at \uni{0335}. There are many
        characters that look like they are precomposed with such an overlay,
        for example <\charis{ɫ~ᵬ~ᵭ~ᵱ}> or <\charis{ƚ~ɨ~ɉ~ɍ}>, or also the
        example of <ø> given at the start of this section. However, they are 
        not decomposed in NFD normalization.
  \item Second, the same situation also occurs with combining characters
        designated as ``combining hook'', like < {\large ̡}> \textsc{combining
        palatalized hook below} at \uni{0321}. This element seems to occur in
        precomposed characters like <\charis{ᶀ~ᶁ~ᶂ~ᶄ}>. However, they are 
        not decomposed in NFD normalization.

\end{itemize}

To harmonize the encoding in these cases it is not sufficient to use Unicode 
normalization. Additional checks are necessary, for example by using orthography 
profiles (see Chapter~\ref{orthography-profiles}).
